---
title: "Defining Statistical Distributions Based on Functional Traits"
author:
  - name: "Rebecca Fisher"
    orcid: "0000-0001-5148-6731"
    affiliations: [aims, iomrc]
  - name: "Diego Barneche"
    corresponding: true
    email: "D.Barneche@aims.gov.au"
    orcid: "0000-0002-4568-2362"
    affiliations: [aims, iomrc]
  - name: "Katherine Cure"
    orcid: "0000-0002-2922-6391"
    affiliations: [aims, iomrc]
  - name: "Matthew Birt"
    orcid: "0000-0003-2233-0716"
    affiliations: [aims, iomrc]
affiliations:
  - id: aims
    name: "Australian Institute of Marine Science, Australia"
  - id: iomrc
    name: "Oceans Institute, The University of Western Australia, Australia"
title-block-published: "Last updated"  
date: now
date-format: long
format:
  pdf:
    number-depth: 2
    documentclass: article
#    classoption: fleqn
    geometry: margin=1in
    papersize: a4
    pdf-engine: xelatex
    keep-tex: true
    toc: false
    math:
      number: true      # Enable equation numbering
      labels: all       # Number all equations
      section: false    # Don't use section numbers (e.g., just (1), (2), ...)
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

# Introduction


# Methods


```{r}
#| label: example
#| eval: false


                 ci=TRUE, ci_method = c("weighted_samples", "multi_fixed"), nboot=10)

```

# Results

<!-- Here's some example analysis code: -->

```{r}
#| label: get-data
#| eval: false
# Note the path that we need to use to access our data files when rendering this document

```

```{r}
#| label: fig-demo-plot
#| fig-cap: "A plot of random numbers"
plot(rnorm(10))
```

@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`

```{r}
#| label: demo-inline-code
#| echo: false
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
